---
title: "Income tax calculator table for labor contracts in Ecuador"
subtitle: "Do you work or plan to work in Ecuador under a dependency relationship? Calculate your tax burden with this tool."
author: 
  - Alex Bajaña
  - Carla Chamorro 
date: November, 2022
title-block-banner: true
format: 
  html:
    css: css/new_css.css
    backgroundcolor: "#03071e"
    fontcolor: white
    page-layout: full
    reference-location: document
server: shiny
execute:
  echo: false
  warning: false
project:
  type: website
---


```{r setup}

#| include: false 

#  Libraries:

library(shiny)
library(tidyverse)
library(gt)
library(shinyBS)
library(tippy)
library(scales)
library(ggrepel)
library(readxl)
library(gtExtras)
library(sparkline)
library(plotly)
library(shinyWidgets)
library(pins)

# File with the exempt income brackets

franjas <- read_excel("data/franjas.xlsx")

# Custom theme for the tables in the app:

table_theme <- function(data, ...){
  data %>% 
    tab_options(
      table.background.color = "#03071e",
      column_labels.background.color = "#3b499c",
      row.striping.background_color = "#051052",
      ...
    ) %>% 
    opt_row_striping()
}


```

## Key concepts

<div class="row">

::: column
### What is income tax?

------------------------------------------------------------------------

Income tax[^1] is levied on income obtained by individuals, undivided inheritances[^2]  and corporations, whether domestic or foreign. The tax year is from January 1 to December 31 (National Assembly of Ecuador, 2004; Presidency of the Republic of Ecuador, 2010).

Individuals, whether they are dependent or not, and whether or not they are obliged[^3] to keep accounting records, are obliged to file their annual income tax form.

:::



::: column
### Labor under a relationship of dependency 

------------------------------------------------------------------------

Article 8 of the Labor Code defines an individual contract as: "(...) an agreement by virtue of which a person undertakes to render lawful and personal services to another or others, under his or her dependence, for a remuneration fixed by agreement, law, collective contract or custom" (National Congress, 2005).

Based on this article, the condition of a dependent relationship refers to the fact that the worker maintains a relationship of subordination to his employer and obedience to the provisions for the development of labor activities.

:::

<br>

------------------------------------------------------------------------

## Income tax calculation for 2022

**The example is developed for a person who earns 3 unified basic salary (SBU 2022) per month, that is, 1275 USD, if you want to generate your own calculation, click on the button**

```{r}

# Init button:

actionButton(inputId = "datos",
             label = "Enter your data",
             width = "100%")
```

<br>


```{r modal-events}
#| context: server

# Launch the modal dialog with all the inputs:

observeEvent(input$datos,{
  showModal(
    modalDialog(
      h2("Tell us about you:"),
      HTML("To declare your income and expenses consider monthly values."),
      hr(),
      with_tippy(
        numericInput(inputId = "ingreso_lab", 
                     label =   div(icon("money-bill-wheat", style = "color:white;"),
                                   HTML("	&nbsp;	&nbsp;	&nbsp;¿What is your monthly wage?")),
                     min = 1,
                     max = 999999,
                     value = 472),
        tooltip = "It only considers income under a dependency relationship. Your monthly salary corresponds to your net income minus your contributions to the IESS (Social security).", 
        placement = "right"
      ),
      hr(),
      with_tippy(
        numericInput(inputId = "ingreso_extra", 
                     label =  div(icon("money-bill-trend-up", style = "color:white;"),
                                  HTML("	&nbsp;	&nbsp;	&nbsp;Do you receive other income such as: utilities, commissions, overtime or other bonuses?")),
                     min = 1,
                     max = 999999,
                     value = 500),
        tooltip = "This income will be added to your total employment income. Consider an annual estimate.", 
        placement = "right"
      ),
      hr(),
      with_tippy(
        sliderInput(inputId = "meses",
                    label = div(icon("calendar-days", style = "color:white;"),
                                HTML("	&nbsp;	&nbsp;	&nbsp;Number of months worked:")),
                    min = 1,
                    max = 12,
                    value = 12),
        tooltip = "Number of months worked with the same employer", 
        placement = "right"),
      h2("Declaration of personal expenses"),
      HTML("To declare your income and expenses consider estimated annual values."),
      hr(),
      with_tippy(
        numericInput(inputId = "gasto_vivienda", 
                     label =  div(icon("house", style = "color:white;"),
                                  HTML("	&nbsp;	&nbsp;	&nbsp;Housing")),
                     min = 1,
                     max = 999999,
                     value = 472),
        tooltip = HTML("<ul>
                         <li>Rent
                         <li>Mortgage loan interest
                         <li>Basic services
                         <li>Property tax
                         <li>Alimony
                         <li>Other expenses for the acquisition, construction, remodeling, expansion, improvement and maintenance of a single real estate property.
                         </ul>
In all the cases indicated in this item and in the previous ones, the expenses will be considered by reason of a single dwelling.."), 
        placement = "right"),
      hr(),
      with_tippy(
        numericInput(inputId = "gasto_educa", 
               label =  div(icon("school", style = "color:white;"),
                            HTML("	&nbsp;	&nbsp;	&nbsp;Education, art and culture")),
               min = 1,
               max = 999999,
               value = 472),
        tooltip = HTML("<ul>
                         <li>Tuition and pension
                         <li>School supplies and textbooks
                         <li>Education for people with disabilities
                         <li>Child care and/or development
                         <li>Uniforms
                         <li>Living and performing arts, literary and narrative, cinematographic and audiovisual, musical and sound
                         <li>Promotion and dissemination of social memory and heritage
                         <li>Crafts
                         <li>Alimony
                         </ul>"), 
        placement = "right"),
hr(),
    with_tippy(
      numericInput(inputId = "gasto_salud", 
                   label =  div(icon("notes-medical", style = "color:white;"),
                                HTML("	&nbsp;	&nbsp;	&nbsp;Health")),
                   min = 1,
                   max = 999999,
                   value = 472),
      tooltip = HTML("<ul>
                         <li>Health professional fees
                         <li>Health services
                         <li>Medications, medical supplies, lenses, orthoses, prostheses and other health accessories
                         <li>Prepaid medicine and health insurance premium
                         <li>Unreimbursed deductible from private insurance settlement
                         <li>Alimony
                         <li>Other expenses: Related to physical and mental well-being, as well as those for prevention, recovery and rehabilitation of health
                         </ul>"), 
      placement = "right"),
hr(),
with_tippy(
  numericInput(inputId = "gasto_alimen", 
               label =  div(icon("utensils", style = "color:white;"),
                            HTML("	&nbsp;	&nbsp;	&nbsp;Nourishment")),
               min = 1,
               max = 999999,
               value = 472),
  tooltip = HTML("<ul>
                          <li>Purchases of food for human consumption and other natural or artificial products that human beings eat to survive or for nutrition
                          <li>Alimony, duly supported by a mediation act or court decision
                          <li>Purchase of food in prepared food expense centers.
                          </ul>"), 
  placement = "right"),
hr(),
with_tippy(
  numericInput(inputId = "gasto_vestimenta", 
               label =  div(icon("shirt", style = "color:white;"),
                            HTML("	&nbsp;	&nbsp;	&nbsp;Clothing")),
               min = 1,
               max = 999999,
               value = 472),
  tooltip = HTML("<ul>
                         <li>Clothing expenses will be considered those made for any type of clothing
                         <li>Alimony, duly supported by a mediation record or judicial resolution
                         </ul>"), 
  placement = "right"),
hr(),
with_tippy(
  numericInput(inputId = "gasto_turismo", 
               label =  div(icon("compass", style = "color:white;"),
                            HTML("	&nbsp;	&nbsp;	&nbsp;Turism")),
               min = 1,
               max = 999999,
               value = 472),
  tooltip = HTML("<ul>
                         <li>Accommodation
                         <li>Transportation
                         <li>Food and beverage services
                         <li>Tourist operation
                         <li>Tourist intermediation
                         <li>Amusement parks
                         <li>Alimony
                         </ul>"), 
  placement = "right"),

footer = tagList(
  actionButton(inputId = "procesar",
               label = "Continue",width = "100%") ,
  size = "l"
), style = "overflow-x: auto;"
    )
  )
})

# Execute the calculations:

observeEvent(input$procesar,{
  
  tabla_presentar$presentar <- tibble(
    ingresos = c(input$ingreso_lab*input$meses + input$ingreso_extra,
                 NA_real_,NA_real_,NA_real_,NA_real_,NA_real_),
    gastos_clase = c("Housing",
                     "Education, art and culture",
                     "Health",
                     "Nourishment",
                     "Clothing",
                     "Turism"),
    gastos_valor = c(input$gasto_vivienda,
                     input$gasto_educa,
                     input$gasto_salud,
                     input$gasto_alimen,
                     input$gasto_vestimenta,
                     input$gasto_turismo)
  )
  
  removeModal()
})

# Glosaries modals ------------------------------------------------

# Expenses glosary:

observeEvent(input$glo_gas,{
             showModal(
               modalDialog(
                          includeHTML("glosario_gastos.html"),
                 footer = actionButton(inputId = "close_glo_gas","Close"),
                 size = "l", style = "overflow-wrap: break-word;"
               )
             )
  })

observeEvent(input$close_glo_gas,{
  removeModal()
})

# Income glosary:

observeEvent(input$glo_ing,{
             showModal(
               modalDialog(
                          includeHTML("glosario_ingresos.html"),
                 footer = actionButton(inputId = "close_glo_ing","Close"),
                 size = "l", style = "overflow-wrap: break-word;"
               )
             )
  })

observeEvent(input$close_glo_ing,{
  removeModal()
})

# Taxable income glosary:

observeEvent(input$glo_bas,{
             showModal(
               modalDialog(
                          includeHTML("glosario_bases.html"),
                 footer = actionButton(inputId = "close_glo_bas","Close"),
                 size = "l", style = "overflow-wrap: break-word;"
               )
             )
  })

observeEvent(input$close_glo_bas,{
  removeModal()
})


```

## Inputs for calculating income tax

In this table, find a summary of your personal income and expenses.

Additionally, we present you with a **tax summary**, which indicates the parameters to consider in your calculation.

```{r}
#| context: server


# Reactive values ----------------------------------------------

# Seed table:

tabla_presentar <- reactiveValues(
  presentar = tibble(
    ingresos = c(425*3*12,NA_real_,NA_real_,NA_real_,NA_real_,NA_real_),
    gastos_clase = c("Housing",
                     "Education, art and culture",
                     "Health",
                     "Nourishment",
                     "Clothing",
                     "Turism"),
     gastos_valor = c( 3000, # input$gasto_vivienda,
                       2160, # input$gasto_educa,
                       1800, # input$gasto_salud,
                       2640, # input$gasto_alimen,
                       1200, # input$gasto_vestimenta,
                       800   # input$gasto_turismo
       )
    
  )
)

# Income and proportions of expenses:

tabla_ingresos <- reactive({
  
  tabla_presentar$presentar  %>% 
    mutate(fraccion = gastos_valor/sum(ingresos,na.rm = T),
           total = sum(ingresos,na.rm = T),
           saldo = total - sum(gastos_valor,na.rm = T),
           otros = sum(gastos_valor,na.rm = T) - gastos_valor,
           gasto_copia = gastos_valor) %>% 
    mutate(plot = NA_character_)
  
})

# Plots for the bar of expenses:

plots_barras <- reactive({
  
  max_gasto <- tabla_ingresos() %>% 
    pull(gastos_valor) %>% 
    max(.,na.rm = T)
  
 plots <- tabla_ingresos() %>% 
   split(.$gastos_clase) %>% 
   map(
     ~.x %>% 
       ggplot() +
       geom_col(aes(x = gastos_valor,y = "a"))+
       xlim(c(0,max_gasto))+
       theme_void() +
       theme(axis.line = element_line(color = "white",linewidth = 3))
   )
 
 plots[c("Housing",
                     "Education, art and culture",
                     "Health",
                     "Nourishment",
                     "Clothing",
                     "Turism")]
  
})

# Get the parameters for the IR calculation:

tabla_resumen_impositivo <- reactive({
  
  income <- tabla_ingresos() %>% 
    pull(ingresos) %>% 
    reduce(sum,na.rm = T)
  
  franjas %>% 
  mutate(comparacion = income,
         across(.cols = c(fraccion_basica,exceso_hasta,impuesto_fraccion ),as.numeric),
         regla = comparacion > fraccion_basica & comparacion < exceso_hasta,
         cuota = if_else(income > 24090.3,0.1,0.2)) %>% 
  filter(regla) %>% 
  select(fraccion_basica,
         impuesto_fraccion,
         impuesto_excedente,
         cuota)
})

# Reshape the parameters:

resumen_impositivo_long <- reactive({
  
  tabla_resumen_impositivo() %>% 
    mutate(across(c(fraccion_basica,
                    impuesto_fraccion),
                  scales::dollar),
           across(c(impuesto_excedente,
                    cuota),
                  scales::percent)) %>% 
    pivot_longer(cols = everything()) %>% 
    mutate(name = case_when(
      name == "fraccion_basica" ~ "Basic fraction (c)",
      name == "impuesto_fraccion" ~ "Tax over the basic fraction (d)",
      name == "impuesto_excedente" ~ "Tariff over the excess (e)",
      name == "cuota" ~ "Deductible fee (f)"
    )) %>% 
    bind_rows(tibble(name = c(NA_character_,NA_character_),
                     value = c(NA_character_,NA_character_)))
  
})

# Income tax calculation:

calculo_impuesto <- reactive({
  
tabla_ingresos() %>% 
    summarise(across(c(ingresos, gastos_valor),sum,na.rm = T)) %>% 
    mutate(gastos_valor = pmin(gastos_valor,754.17*7)) %>% 
    cbind(tabla_resumen_impositivo()) %>% 
    transmute(preliminar_1 = ingresos-fraccion_basica,
              preliminar_2 = cuota*gastos_valor,
              preliminar_3 = preliminar_1*impuesto_excedente,
              total = impuesto_fraccion + preliminar_3 - preliminar_2,
              total = if_else(total < 0,0,total),
              tie = total/ingresos) 

})

# Deducible expenses message:

output$expenses <- renderUI({
  
  gastos <- tabla_ingresos() %>% 
    summarise(across(c(ingresos, gastos_valor),sum,na.rm = T)) %>% 
    pull(gastos_valor)
  
  
  if(gastos > 754.17*7){
    
    HTML("Your expenses <b>exceed</b> the legal deductible amount equivalent to 7 basic family baskets, that is, 5279.19 USD. <b>This value is applied as your deduction for personal expenses.</b><br><br>")
    
  }else{
    
    HTML("Your expenses are deductible according to the fee presented in the table.")
    
  }
  
})

# Outputs -------------------------------------------------

# Tax calculation arguments:

output$tabla_argumentos <- render_gt({
  
  
  bind_cols(tabla_ingresos(), 
            resumen_impositivo_long())  %>% 
    select(ingresos,
           gastos_clase,
           gastos_valor,
           plot,
           fraccion,
           name,
           value) %>% 
    gt() %>% 
    
    # Column formatting:
    
    text_transform(
      locations = cells_body(plot),
      fn = function(x) {
        
        map(plots_barras(),
            ggplot_image, 
            height = px(30),
            aspect_ratio = 3)
        
      }
    ) %>% 
    fmt_currency(
      columns = c(gastos_valor,ingresos),
      currency = "USD"
    )  %>%
    fmt_percent(
      columns = fraccion
    )  %>%
    sub_missing(
      columns = c(ingresos,
                  name,
                  value),
      missing_text = ""
    ) %>% 
      
    
    # Column labeling:
    
    cols_align(columns = everything(),
               align = "center") %>% 
    cols_label(
      ingresos = md("**Income (a)**"),
      gastos_clase = "",
      gastos_valor = md("**Value**"),
      name = md("**Concept**"),
      value = md("**Value**"),
      plot = "",
      fraccion = md("**% of income**")) %>%
    tab_spanner(label =  md("**Personal expenses (b)**"),
                columns = c(gastos_clase,
                            gastos_valor,
                            plot,
                            fraccion)) %>%
    tab_spanner(label =  md("**tax summary**"),
                columns = c(name,
                            value))  %>%
    
    # Summaries:
    
    grand_summary_rows(
      columns = c(ingresos, gastos_valor),
      fns = list(
        Total = ~sum(.,na.rm = T)),
      formatter = fmt_currency,
      currency = "USD",
      use_seps = FALSE,
      missing_text = ""
    ) %>%
    grand_summary_rows(
      columns = fraccion,
      fns = list(
        Total = ~sum(.,na.rm = T)),
      formatter = fmt_percent,
      missing_text = "",
      use_seps = FALSE
    ) %>%
    
    # Cols with fixed width:
    
    cols_width(ingresos ~ px(200),
               plot ~ px(150)) %>%
    # Dividers:
    
    gt_add_divider(columns = "ingresos", 
                   style = "solid") %>%
    gt_add_divider(columns = "fraccion", 
                   style = "solid") %>%
    
    # Footnote:
    
    # gt::tab_footnote(
    #   footnote = "El valor en paréntesis indica el porcentaje que el gasto anual representa en el total de los ingresos anuales",locations = gt::cells_column_labels(columns = gastos_valor)
    # ) %>% 
    
    # Cells styles:
    
   tab_style(
     style = list(
       cell_borders(
         sides = c("top", "bottom"),
         color = "#051052",
         weight = px(2)
       ),
       cell_borders(
         sides = c("top", "bottom"),
         color = "#051052",
         weight = px(2)
       ),
       cell_borders(
         sides = c("top", "bottom"),
         color = "#051052",
         weight = px(2)
       ),
       cell_borders(
         sides = "bottom",
         color = "#03071e",
         weight = px(2),
       )
     ),
     locations = list(
       cells_body(
         columns = c(ingresos),
         rows = is.na(ingresos)
       ),
       cells_body(
         columns = name,
         rows = is.na(name)
       ),
       cells_body(
         columns = value,
         rows = is.na(value)
       ),
       cells_stub(rows = everything())
     )
  )  %>% 
    
    # Theme:
    
    table_theme() 
  
})

# Income tax calculation table:


output$tabla_resultado <- render_gt({

  calculo_impuesto() %>%
    gt() %>%
    cols_label(
      preliminar_1 = md('<p align="right">Taxable Income (a)<br> - Basic Fraction (c) <hr style=" background-color: #5399EE; height:2pt; margin-bottom:-1px; margin-top:-1px;" /> <b>Taxed Surplus (g)</b></p>'),
      preliminar_2 = md('<p align="right">Total expenses (b)<br> x Deductible installment (f) <hr style=" background-color: #5399EE; height:2pt; margin-bottom:-1px; margin-top:-1px;" /> <b>Expense deduction p. (h)</b></p>'),
      preliminar_3 = md('<p align="right">Taxable income (g)<br> x Excess rate (e) <hr style=" background-color: #5399EE; height:2pt; margin-bottom:-1px; margin-top:-1px;" /> <b>Preliminary tax (i)</b></p>'),
      total = md('<p align="right">Preliminary tax (i)<br> + Tax on fraction (d)<br> - Expense deduction (h) <hr style=" background-color: #5399EE; height :2pt; margin-bottom:-1px; margin-top:-1px;" /><b> Income tax</b></p>'),
      tie = md("**How much does *income tax* represent in my income?**")) %>%
    fmt_currency(
      columns = c(preliminar_1,
                  preliminar_2,
                  preliminar_3,
                  total),
      currency = "USD"
    )  %>%
    fmt_percent(
      columns = tie
    )  %>%
    cols_width(tie ~ px(150))  %>% 
    table_theme()  


})


# Outcome message:

output$mensaje <- renderUI({
  
  impuesto <- calculo_impuesto() %>% 
    pull(total)
  
  if(impuesto == 0){
    
    HTML("<b>!In good time, you should not pay income tax!</b>
           <br>")
    
  }else{
    HTML(str_c("<b>The total income tax</b> that you must pay is ", impuesto,"$ USD"), 
         "<br>")
  }
  
})

```


```{r}

# Present the parameter table:

gt_output("tabla_argumentos")
```


```{r}
uiOutput("expenses")
```

With the inputs presented, we will calculate the income tax:

```{r}

# Present the outcome table:

gt_output("tabla_resultado")

# plotlyOutput("plot_foot",height = "100px")

```

Fiscal obligations:

```{r}

# Message:

uiOutput("mensaje")
```

<br>

Remember that you must present:

* The <b>Annex of personal expenses</b>[^4] in February 2023
* The <b>Income Tax Form </b> in March 2023

Check the ninth digit of your identity card to find out the date on which you must present your annex and declaration.

## + More FYI

If you have any doubts, you can consult the following glossaries:

```{r}

# Buttons for the glosaries:

 actionGroupButtons(
            inputIds = c("glo_gas", "glo_ing", "glo_bas"),
            labels = list("Glosario de gastos",
                          "Glosario de ingresos",
                          "Glosario base imponible"),
            status = "primary",
            fullwidth = TRUE
        )


```

<br>
<br>

For additional information you can contact the authors:

- Alex Bajaña (alexvbr@ergostats.org)
- Carla Chamorro (carlachamorro@gmail.com)

[^1]: Article 2 Organic Law of the Internal Tax Regime and article 1 of the Regulations of the Law.

[^2]: Undivided estates for income tax settlement are considered as independent economic units, that is, they are understood and treated as a natural person. Undivided successions are temporary taxpayers, their existence begins with the death of a natural person and ends by court order or by testament of the deceased, that is, when the heirs incorporate the inheritance of the deceased into their own patrimony (Presidency of the Republic of Ecuador , 2010).

[^3]: Those who meet the following criteria are required to keep accounts: i) \$300,000 dollars of gross income in the previous fiscal year, ii) \$240,000 dollars of costs and expenses in the previous fiscal year, and iii) $180,000 dollars by own capital on January 1, 2022 or at the start of the activity (Internal Revenue Service, 2021).

[^4]: The Personal Expenses Annex must be presented if the deductible personal expenses exceed 50% of the basic fraction deducted from Income Tax of natural persons. People who work in a dependency relationship are not required to submit the Income Tax return as long as they work with a single employer and do not project personal expenses, or if they have done so, there are no values ​​that must be reliquidated.
